---
title: "countries: a **R** package to select and map a subset of countries"
output: pdf_document
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{countries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width="100%",
  fig.width = 10,
  fig.align = 'center',
  message = FALSE,
  warning = FALSE
)
```


**countries** is a small package that performs three tasks:

- constructing a `sf` with a relevant subset of countries,
- joining this `sf` with a tibble,
- providing a simple `plot` method to get a default plot. 

Is is loaded using:


```{r echo = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
```

```{r }
library(countries)
```

and use shape files provided by the [Natural
Earth](https://www.naturalearthdata.com/) website and use extensively
**sf** [@PEBE:18] and **ggplot2** [@WICK:16] and some other **R**
packages^[In particular **ggrepel** [@SLOW:21] for labels and
**classInt** [@BIVA:23] to compute automatically class intervals.].

# What is a country ? 

This is a most difficult question that it seems.^[In @MASS:SOUT:23, a
package that enables to download within **R** maps from naturalheart,
there is a vignette on this question.] There are 193 member states of
the United Nations and 2 general assembly observer states (Holly See
and State of Palestine).  Some countries have dependencies which are
often overseas territories, like for example French Polynesia and New
Caledonia for France, which have a special status somewhere between a
normal region and a sovereign state.  Finally, it is convenient, at
least as far as drawing maps is concerned, to cut some countries in
different pieces, the main territory and some parts. Consider for
example Spain.

```{r echo = FALSE}
sf::st_geometry(countries:::Spain) %>% plot
```
Spain consists of a main, continental territory and two sets of
Islands, the Balearic Islands and the Canaries Islands, which are
Spanish provinces. It's not a problem to plot the Balearic Islands
along with continental Spain as they are almost entirely contained in
the bounding box of Continental Spain. It is not the case for Canaries
Islands which are situated near the coasts of Morocco. Therefore, it
is convenient to consider two different geometries for Spain:
continental Spain and the Balearic on the one hand and the Canaries
Islands on the other hand. This division doesn't obey to any political
rule but is performed only for plotting convenience.
In the same spirit, the United States of America is splitted in three
(Mainland USA, Alaska and Hawaii) and Italy is kept as one geometry as
Sicily and Sardinia are close enough to mainland Italy.

We'll therefore consider three categories of entities:

- main territory of sovereign countries (most of the time the whole
  country),
- parts of a sovereign country,
- dependencies of a sovereign country.

**countries** is based on **Natural Earth**, using the most detailed
scale, which is `1:10,000,000` (1cm = 100km). **Natural Earth**
provides different shape files containing the administrative borders
of countries, with a different number of entities. They are called
**sovereignty**, **countries**, **units** and **subunits**.
For example, in the **sovereignty** file, France is a unique line (and
therefore a unique multipolygon), as in the **countries** file, there
is one line for each dependency (New Caledonia and French Pacific for
example) and one unique line for continental France and the 5 overseas
departments. In the **units** file, each overseas departments has its
own line and in the **subunits** file Corsica also has its own line. 

In the **countries** package, we start from the **countries** file,
which contains 258 entities, and we split some of them to obtain
finally 295 entities, categorized as follow:

- **main** (199): roughly corresponds to the main parts of the
  sovereign countries, ie the 193 UN recognized countries, the 2 UN
  observer countries (Palestine and Vatican), and 4 not or not fully
  recognized countries (Kosovo, Somaliland, Northern Cyprus and
  Taiwan),
- **part**: 37 parts of sovereign countries (including Macao and Hong
  Kong),
- **dependency**: 48 dependencies of sovereign countries,
- **indeterminate** : 11 territories, Western Sahara, Brazilian
  Island, Cyprus No Mans Area, Siachen Glacier, Southern Patagonian
  Ice Field, Bir Tawil, Antarctica, Spratly Islands, Bajo Nuevo Bank
  (Petrel Is.), Serranilla Bank, Scarborough Reef

The raw information about countries is stored in a `sf` called
`ne_countries`^[We use `rmapshaper::ms_simplify` to reduce the size of
the file by a factor of about 10, except for small islands which were
kept as in the initial file.]. The details of the computation are
presented in the last section of this document. Two geometries are
present in this `sf`, `polygon` for the administrative boundaries and
`point` for the point coordinate of the capital:

```{r }
ne_countries %>% as_tibble %>% select(- polygon, - point) %>%
    print(n = 2, width = Inf)
```

Each country is identified either by its name (`country`) and, for 249
of them, by the two and three digits **ISO 3166-1** code (respectively
`iso2` and `iso3`). `type` indicates whether the entity is the main
part of a sovereign country (`"main"`), a part or a dependency of a
sovereign country (`"part"` or "`dependency"`), or an indeterminate
territory (`"indeterminate"`). `sovereign` is the name of the
sovereign country (equal to `country` is the entity is of the `"main"`
category). `capital` is the name of the capital of the country and
`status` is the United Nation status. The name of the country is also
provided in 5 languages in the columns `en`, `fr`, `de`, `es`, `it`.

Countries are grouped in different entities:

- United Nation's regions `region` (Africa, Americas, Asia, Europe,
  Oceania) which are decomposed in 22 subregions `subregion`,
- World Bank's regions `wbregion` (Antarctica, East Asia & Pacific,
  Europe & Central Asia, Latin America & North America, South Asia and
  Sub-Saharan Africa),
- Economic group `economy`,
- Income group `income`.

Finally, two numeric covariates are provided, the population (`pop`)
and the gross domestic product (`gdp`).

`ne_towns` is another `sf` that contains 7342 towns, obtained from a
shape file of **Natural Earth** called populated places; it contains
their names `name`, the iso codes of the country they belong to
(`iso2` and `iso3`), a boolean indicating whether it is a capital
(`capital`) the population `pop` and the point coordinates `point`.

```{r }
ne_towns %>% print(n = 2)
```

These two `sf` are exported by `countries` and can therefore be used
directly. 


# Selecting countries

The `countries` function is provide to extract a subset of
countries. It's first argument is called `name` and its default value
is `NA`. In this case all the countries are returned.

```{r eval = FALSE}
countries()
```

`name` is a vector of character that should contain either countries,
regions or subregions (but not a mixture). Let's first select one
country, for example France:

```{r }
countries("France") %>% as_tibble %>% select(1:5)
```

by default, only the main part of France is returned. The parts
(dependencies) can also be returned by setting the `part`
(`dependency`) argument to `TRUE`:

```{r }
fr_parts <- countries("France", part = TRUE)
fr_parts %>% pull(country)
countries("France", dependency = TRUE) %>% pull(country)
```

`countries` returns an object of class `countries` that inherits from
`sf`. It has a `type` attribute that is either `country`, `region`,
`subregion` or `world`, depending on the kind of entities selected in
the `name` argument. It also have a `bb` attribute which contains the
bounding box and, if `coastlines` is `TRUE` (the default), a `bg`
attribute that is a `sfc` containing the coastlines in the bounding
box of the selected territories. For example, for France with the
parts selected, we have:

```{r }
fr_parts %>% attr("bb") %>% plot(border = "red")
fr_parts %>% attr("bg") %>% plot(add = TRUE)
```

Note that the bounding box for France extends on the south-east to La
Reunion and on the west to Martinique and Guadeloupe.
If a region or a subregion is selected, the `part` and `dependency`
arguments reason on the region / subregion and not on the countries
that belongs to the region / subregion. For example, France is part of
the Western Europ subregion. Selecting this subregion with `part =
TRUE`:

```{r }
countries("Western Europe", part = TRUE) %>% pull(country)
```

doesn't return the overseas departments which are parts of France but
which are not part of the subregion France belongs to.
A `indeterminate` argument, by default equal to `FALSE` enables to
select or not the indeterminate territories.
The `exclude` and `include` arguments are characters which enables to
exclude or include entities from the set selected by the `name`
argument. For example, to get a world map without the Antarctica:

```{r }
countries(exclude = "Antarctica", coastlines = FALSE) %>% plot
```

This is a world map with sovereign countries, without parts and
dependencies. To add Alaska (a part of the USA) and Greenland (a
dependency of Denmark), we use the `include` argument:

```{r }
countries(exclude = "Antarctica", coastlines = FALSE,
          include = c("Alaska", "Greenland")) %>% plot
```

Note the use of the `plot` method for `countries` object. A basic plot
is obtained without argument, we'll develop further the use of this
method latter. Note for now that, contrary to `sf`, the `plot` method
for `countries` object use **ggplot2**, more specifically it is
obtained by a call to `ggplot(x) + geom_sf()`.

# Selecting towns

`countries::towns` function select towns with a first mandatory
argument that can be either a  character vector (as in
`countries::countries`) or a `countries` object:

```{r }
we <- countries(c("France", "Spain"))
towns(we) %>% print(n = 2)
```

is equivalent to:

```{r eval = FALSE}
towns(c("France", "Spain"))
```

`capital` is a boolean that unsures that the capital of every
countries are selected. `size` is a numeric that indicates the minimum
population of the towns to be returned. The default value of `size` is
$0$ so that all the towns of the selected countries are
returned. Consider for example Australia:


```{r }
towns("Australia", size = 2E06)
```
returns the two largest towns of Australia (Melbourne and Sydney). 
Setting `capital` to `TRUE`, we get:

```{r }
towns("Australia", size = 2E06, capital = TRUE)
```

ie, Canberra, which is only 327,700 inhabitants, is returned
because it is the capital of Australia.

Towns can be selected within the `countries::countries` function by
using the `capital` and the `towns` arguments. `capital` has exactly
the same meaning as previously, `towns` can either be a boolean
(`FALSE` no towns, except eventually the capital are returned or
`TRUE`, all the towns are returned) or a numeric which is passed to
the `size` argument of `countries::towns`.

```{r }
aus <- countries("Australia", towns = 2E06, capital = TRUE)
```

When arguments `capital` and/or `towns` are used in
`countries::countries`, the returned `countries` object has a `towns`
attribute that contains the towns:

```{r }
attr(aus, "towns")
```

A `labels` method for `countries`'s objects is provided. It returns a
`sf` with a `POINT` `sfc` which contains the entities that can be
labelled, which can be either countries, capitals, towns or a
combination of them. For example:

```{r }
countries(c("Portugal", "Spain"), towns = 1E06, capital = TRUE) %>%
    labels(var = c("country", "towns", "capital"))
```
The `type` and `name` columns contains entities' types (country, capital or town)
and names. The geometry is the coordinates of the cities for `capital`
and `town` entities and the point obtained by the
`sf::st_point_on_surface` function for countries. We'll see later that
this function can be used to display labels on maps.

# Coordinate reference system

**natural earth** use geographical coordinates with the **WGS84**
datum, the well known **crs** 
``r sf::st_crs(ne_countries)$proj4string``  with **epsg** code 4326.

A common problem is that some territories are on both sides of the 180
E/W longitude. In this case, the `shift` argument can be used, so that
the center of the world map is the 180 longitude and not the Greenwich
(0) longitude (internally, the `sf::st_shift_longitude` function is
used). Consider for example Russia:

```{r }
countries("Russia", coastlines = FALSE) %>% plot
```

A small part of Russia appears on the left hand size of the map. Using
`shift = TRUE`, we get:

```{r }
countries("Russia", coastlines = FALSE, shift = TRUE) %>% plot
```

To use projected **crs**, either the `utm` or the `crs` arguments can
be used. The `crs` argument, if used, is passed to `sf::st_transform`
in order to transform the geometry in the required `crs`. The `utm`
argument enables the use of the Universal Transverse Mercator
coordinate system which divides earth into 60 zones. If `TRUE`, the
most relevant zone is automatically selected but the user can also
select a specific zone using an integer from `0L` to `60L`. For
example, to get a utm projected map of Europe without Russia but with
Turkey and Cyprus:

```{r }
countries("Europe", utm = TRUE, extend = 1.1,
          include = c("Turkey", "Cyprus", "Northern Cyprus"),
          exclude = "Russia") %>% plot
```

The Lambert conform conic projection for Europe can be used by setting
the `crs` argument to the **epsg** code of this **crs** which is 3034:

```{r }
countries("Europe", crs = 3034, extend = 1.5,
          include = c("Turkey", "Cyprus", "Northern Cyprus"),
          exclude = "Russia") %>% plot
```

# Thematic maps

Until now, we drew very basic maps, in order to show the set of
countries selected. More advanced maps can be produced by:

- filling countries using a categorical or a numerical variable, 
- maping the shape or the size of points (that can be either the
capital or the centroid of countries) with a categorical or a
numerical variable,
- adding labels for countries, capitals and/or towns.


For example the `economy` variable is a factor that contains the
economic group (developed, emerging, etc.). To fill the different
countries with colors associated to the modalities of this factor, we
use the `fill` argument:

```{r }
countries("Asia", exclude = "Russia") %>% plot(fill = "economy")
```

The palettes used are from **ColorBrewer**. Any qualitative palettes
can be used using the `palette` argument. For exemple, to use the
`"Dark2"` palette:

```{r eval = FALSE}
countries(c("Asia"), exclude = "Russia") %>%
    plot(fill = "economy", palette = "Dark2")
```

For numeric variables, bins can be constructed using the `bks`
argument. For example, `pop` contains the population of the
countries:

```{r }
countries("Europe", exclude = "Russia") %>%
    plot(fill = "pop", bks = c(0, 1E06, 5E06, 1E07, 5E07, 1E08, Inf))
```

By default, the `"Blues"` palette is used. As previously, any sequential or divergent
palette can be used using the `palette` argument. For example, to use the `"PuBu"` divergent
palette:

```{r eval = FALSE}
countries("Europe", exclude = "Russia") %>%
    plot(fill = "pop", bks = c(0, 1E06, 5E06, 1E07, 5E07, 1E08, Inf),
         palette = "PuOr")
```

The bins can also automatically be computed using the **classInt**
package. For this purpose, the `plot` method have `style` and `n`
arguments that are passed to `classInt::classIntervals`:

```{r }
countries("Europe", exclude = "Russia") %>%
    plot(fill = "pop", n = 10, style = "pretty",
         palette = "Oranges")
```
Points can be added to the map if the `countries` includes a `"towns"`
attribute, which is the case if `capital` and `towns` are not `FALSE`
in the call to the `countries` function. 
If `capital = TRUE` and `towns = FALSE`, the capitals are represented
by a point with a size related to their populations. If `towns` is not
`FALSE`, some non-capital towns are added, with a special shape and a
size related to their population:

```{r }
countries("Europe", exclude = "Russia", capital = TRUE, towns = 1E06) %>%
    plot(fill = "pop", n = 10, style = "pretty",
         palette = "Oranges")
```

One point for each country (which can be either the capital or the
centroid of the country) can also be associated to a numeric or a
categorical variable. For example, we can map the point size of the
centroid to the gross domestic product using the `centroid` argument:

```{r }
countries("Europe", exclude = "Russia") %>%
    plot(fill = "pop", centroid = "gdp", n = 10, style = "pretty",
         palette = "Oranges")
```

and we can perform the same operation (this time for a categorical
variable `income`) for the points defined by the position of the
capital using the `capital` argument, so that the shape of the points
is maped to the categorical variable.

```{r }
countries("Europe", exclude = "Russia", capital = TRUE) %>%
    plot(fill = "pop", capital = "income", n = 10, style = "pretty",
         palette = "Oranges")
```

Labels can be added using the `labels` argument, which is a character
that countains any combination of `"country"`, `"capital"` and
`"towns"`. Moreover, for the names of the countries, 5 different
languages are available. Here, we use the `lang` argument of
`countries::countries` to display countries' names in Spanish:

```{r fig.height = 10, output.width = "70%"}
countries("Europe", exclude = "Russia", capital = TRUE, lang = "es") %>%
    plot(fill = "pop", capital = "income", n = 10, style = "pretty",
         palette = "Oranges", labels = "country") +
    labs(x = NULL, y = NULL) +
    guides(fill = "none", shape = "none")
```

In a small territory, we can also add labels for towns and
capitals. the **ggrepel** package is used to avoid overplotting of
labels. In the next map, we use `name = "Yougoslavia"`, which selects
all the countries of the former Yougloslavia^[`name` can also be set
to `"USSR"` to get all the countries of the former Soviet Union.]:

```{r fig.height = 10, output.width = "70%"}
countries("Yougoslavia", capital = TRUE, towns = TRUE) %>%
    plot(fill = "pop", capital = "income", n = 4, style = "pretty",
         palette = "Oranges", labels = c("country", "capital", "towns")) +
    labs(x = NULL, y = NULL) +
    guides(fill = "none", shape = "none")
```

# External data

The `sf` returned by `countries` can also be joined with an external
tibble. For this purpose, a `left_join` method for `countries`'s
objects is provided. The external tibble must have a column that
identifies the entity, which can be either the name or the 2 or 3
digits iso code. To illustrate this feature, we consider two real
world examples. The first tibble is called `slave_trade` and is used
in @NUNN:08. In this article, the long-term effects of slave trade on
African countries' economic activity is analysed. The data set
contains 52 African countries and the main covariate is the number of
slaves exported from each country divided by the average population
during the slave trade period. We compute this covariate and select
some existing columns, `gdp` (gdp per capita in 2000) and `colony` (a
factor containing the previous colonizator):

```{r }
slave_trade <- slave_trade %>%
    mutate(slaves = slaves / pop) %>%
    select(country, slaves, gdp, colony)
```

The `left_join` method takes only three arguments: the `countries`
object, the external tibble and the column that countains countries'
identifiers. A `check_join` function is provided, with a further
argument called `side`, that checks:

- whether all the countries of the tibble are present in the
  `countries` object (`side = "right"`, the default),
- whether all the countries of the `countries` object are present on
the tibble (`side = "left"`),
- whether the two sets of countries are the same (`side = "both"`).

```{r eval = TRUE}
countries("Africa") %>%
    check_join(slave_trade, by = "country", side = "both")
```

4 countries from the `slave_trade` tibble don't have correspondance
with the `countries` object because of different spellings. On the
contrary, 3 countries of the `countries` object are not present in
`slave_trade`: Eritrea and South Sudan (which were not or freshly
independent by the time of the article), and Somaliland. We then
correct the spelling of the 4 countries in the `slave_trade` tibble
before joining:


```{r eval = TRUE}
slave_trade <- slave_trade %>%
    mutate(country = case_when(country == "Democratic Republic of Congo" ~ "D.R. Congo",
                               country == "Cape Verde Islands" ~ "Cabo Verde",
                               country == "Sao Tome & Principe" ~ "Sao Tome and Principe",
                               country == "Swaziland" ~ "eSwatini",
                               .default = country))
```

and we can now perform the join:

```{r eval = TRUE}
strade <- countries("Africa", capital = TRUE) %>% select(iso2:status, point) %>% 
    left_join(slave_trade, "country")
strade %>% plot(fill = "slaves", n = 5, type = "pretty", capital = "gdp")
```

`sp_solow` contains the data used by @ERTU:KOCH:07 to estimate a
Solow's growth model with externalities and taking into account
spatial dependence.

```{r }
sp_solow %>% print(n = 3)
```

The values of the gross domestic product are given for the years 1960
and 1995, we compute the average growth rate for the period:

```{r }
sp_solow <- sp_solow %>%
    mutate(growth = (gdp95 / gdp60) ^ (1 / 35) - 1)
```
Either `name` or `code` can be used to joint `sp_solow` with the
`countries`' object. It is much safer to use `code` as it avoids the
problem of small differences in countries' names. As a lot of
countries of the world are not present in `sp_solow` (especially most
of the communist countries), we just check whether all the the
countries of `sp_solow` are present in the `countries` object:

```{r eval = TRUE}
countries() %>% check_join(sp_solow, by = "code")
```

The two problems is that the D.R. Congo (`iso3` code `COD`) used to be
called Zaire (`iso3` code `ZAR`) and that Hong Kong, which is a part
of a China in `ne_countries` was considered as a sovereign country by
the time of the study.

```{r eval = TRUE}
sp_solow <- sp_solow %>% mutate(code = ifelse(code == "ZAR", "COD", code))
sps <- countries(include = "Hong Kong", exclude = "Antarctica") %>%
    select(iso2:status, point) %>% 
    left_join(sp_solow, by = "code")
```

We then draw a world map with the color of the countries related to
the annual growth during the 1960-95 period and a point with a size
related to the initial (1960) gdp.

```{r eval = TRUE}
sps %>% plot(fill = "growth", centroid = "gdp60")
```

# Details of the computation

The most aggregate file is called **sovereignty** and
contains 209 entities: the 193 UN countries and 16 more territories
characterized as follow:

- **disputed**: Kosovo,
- **sovereign country**: Northern Cyprus, Somaliland, Taiwan,
  Vatican
- **indeterminate**: Antarctica, Bajo Nuevo Bank (Petrel Is.) (small
  inhabited islands Colombia / United States of America / Nicaragua
  and Jamaica), Bir Tawil (Egypt / Sudan), Brazilian Island (river
  islands Brazil / Uruguay), Cyprus No Mans Area, Scarborough Reef
  (islands China / Philippines / Taiwan), Serranilla Bank (caribbean
  islands Colombia / Nicaragua), Siachen Glacier (India / Pakistan),
  Southern Patagonian Ice Field (Argentina / Chile), Spratly Islands
  (Brunei / China / Malaisia / Philippines / Taiwan / Viet Nam),
  Western Sahara (Morocco / Sahrawi Arab Democratic Republic)

Four more sovereign countries but with not full recognization are
added: Kosovo, Northern Cyprus, Somaliland and Taiwan. Note that
Palestine, one of the two observer states is not included in this
file.

The **countries** shape file contains 49 more entities that are mostly
countries' dependencies. The 258 entities are categorized the
following way:

- **sovereign country** : 185 (**181** UN members + Vatican, Somaliland,
  Taiwan and Northern Cyprus)
- **sovereignty** : **2** (Cuba and Kazakhstan)
- **country** : 19 (including **9** Sovereign countries: Netherlands,
  France, China, Finland, United Kingdom, United States of America,
  Australia, New Zealand, Denmark) + Sint Maarten, Hong Kong, Greenland,
  Curaçao, Aruba, Jersey, Guernsey, Isle of Man, Aland, Macao.
- **disputed** : **1** UN member Israel, Kosovo, Gibraltar, British Indian
Ocean Territory, Falkland Islands,
- **indeterminate** : same 11 entities as previously + Palestine
- **lease** : US Naval Base Guantanamo Bay, Baykonur Cosmodrome
- **dependency** : 33 territories

With a enlarged definition of sovereign countries, we then have $181 +
2 + 9 + 1 = 193$ UN members, 2 observers (Vatican, Palestine) and 4
not fully recognized countries (Kosovo, Nothern Cyprus, Somaliland,
Taiwan), 199 in total.

The dependency set consists of the 33 territories categorized as
such, the remaining 10 territories of the country category, the 3
territories belonging to the disputed category (except Israel and
Kosovo) and the 2 territories in the lease category (Guantanamo and
Baykonur), 48 in total, presented in the following list, with the name
of the sovereign country in bold.

```{r echo = FALSE, results = 'asis'}
deps <- anti_join(countries:::countries_list, countries:::sovereignty_list, by = "country") %>%
    select(country, sovereign) %>%
    filter(sovereign != "Israel") %>% 
    tidyr::nest(.by = sovereign)
names_deps <- deps[[1]]
deps <- purrr::map(deps[[2]], ~ paste(.x[[1]], collapse = ", "))
names(deps) <- names_deps
for (i in 1:length(deps)){
    sov <- names(deps)[i]
    adep <- deps[i]
    cat(paste("- **", sov, "**: ", adep, "\n", sep = ""))
}
```

The 11 remaining territories correspond to the Indeterminate
category.

As stated previously, there is a need to split some entities in
different features. The complete list of the 39 part territories is
given below:

- **France (5)**: Reunion, Mayotte, Guyane, Martinique, Guadeloupe (the 5
  overseas deparments),
- **United States of America (2)**: Alaska and Haiwaii
- **Norway (2)**: Bouvet (a small Island in Antarctica) and Svalbard and Jan
  Mayen (a set of Islands far north of the Norwegian coasts),
- **Netherlands (1)**: Bonaire, Sint Eustatius and Saba (Caribbean Islands)
- **Portugal (2)**: Azores and Madeira
- **Spain (1)**: Cannaries
- **Mauritius (2)**: Rodrigues, Agalega
- **New Zealand (6)**: Tokelau, Chatman, Kermadec, Auckland, Campbelle,
  Antipodes
- **Chile (1)**: Easter Island
- **Colombia (1)**: San Andres, Providencia and Santa Catalina
- **South Africa (1)**: Prince Edward Islands
- **Ecuador (1)**: Galapagos
- **Australia (1)**: Macquarie Island
- **Denmark (1)**: Bornholm
- **Equatorial Guinea (1)**: Annobon
- **Seychelles (6)**: Aldabra, Coetivy, Alphonse, Attol Saint-Joseph, Platte,
  denis
- **Antigua and Barbuda (1)**: Redonda
- **Indian Ocean Territories (2)**: Christmas Islands, Cocos Islands
- **United States Minor outlying Islands (2)**: Navassa (Navassa is a small
  caribbean islands as the other ones are in the Pacific)

<!-- Initiallement : -->

<!-- main 199 -->
<!-- dep 48  -->
<!-- misc 11 -->
<!-- total 258 -->

<!-- on ajoute 39, dont deux splits -->
<!-- main 199 -->
<!-- dep 48 - 2 = 46 -->
<!-- misc 11 -->
<!-- part 39 -->

<!-- on transfere HK et Macao de dep à parts -->
<!-- on transfere USMOI x 2 et UIT x 2 de parts à dep -->
<!-- main 199 -->
<!-- dep 46 - 2 + 4 = 48 -->
<!-- part 39 - 4 + 2 = 37 -->
<!-- misc 11 -->

Some very small islands from Japan, Brazil, the United Kingdom and
Venezuela were removed.

Note that the Indian Ocean Territories and the United States Minor
outlying Islands are splited in two. We then have 295 entities, 258
from the **countries** file minus 2 (Indian Ocean Territories and US
minor outlying islands) plus 39, grouped in 4 categories, as
described in the first section of this document.

# References

